---
title: "Drug Consumption"
author: "Nhat Bui, Johan Ferreira, Thilo Holstein"
date: "2025-03-06"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
# Introduction

Drug use is a significant risk behavior with serious health consequences for individuals and society. Multiple factors contribute to initial drug use, including psychological, social, individual, environmental, and economic elements, as well as personality traits. While legal substances like sugar, alcohol, and tobacco cause more premature deaths, illegal recreational drugs still create substantial social and personal problems.

In this data science project, we aim to identify factors and patterns potentially explaining drug use behaviors through machine learning techniques. By analyzing demographic, psychological, and social variables in our dataset, we'll aim to uncover potential predictors, use machine learning methods to understand the complex relationships surrounding drug consumption, demonstrating how machine learning can reveal insights into behavioral patterns. While our findings won't directly inform interventions, this project showcases how data-driven approaches can enhance our understanding of complex social phenomena and provide valuable practice in applying machine learning to real-world datasets.

The database contains records for 1,885 respondents with 12 attributes including personality measurements (NEO-FFI-R, BIS-11, ImpSS), demographics (education, age, gender, country, ethnicity), and self-reported usage of 18 drugs plus one fictitious drug (Semeron). Drug use is classified into seven categories ranging from "Never Used" to "Used in Last Day." All input attributes are quantified as real values, creating 18 distinct classification problems corresponding to each drug. A detailed description of the variables can be found in the Column Decsription text file.

# Cleaning and formatting the dataset

```{r echo=FALSE, cache=TRUE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(reshape2)
library(kableExtra)
library(knitr)
library(gridExtra)
library(tidyr)
```

```{r echo=FALSE, cache=TRUE}
# Read the CSV file
drug_data <- read.csv("Data/drug_consumption.csv")
```

## Fomatting the dataset

The original data set had all the values for most of the variables set to a random floating number representing a specific categorical value, we believe this was done in order to remove bias from the dataset. As the requirements of this project is different form the data set's original intention we had to replace these values with the original values in order to complete all the required steps for our project.

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Define mappings and column information
###############################################################################

# Column names for the dataset
column_names <- c(
  "Index", "ID", "Age", "Gender", "Education", "Country", "Ethnicity", 
  "Nscore", "Escore", "Oscore", "Ascore", "Cscore", "Impulsive", "SS", 
  "Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", "Choc", "Coke", 
  "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh", "LSD", "Meth", 
  "Mushrooms", "Nicotine", "Semer", "VSA"
)

# Drug column names
drug_columns <- c(
  "Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", 
  "Choc", "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", 
  "Legalh", "LSD", "Meth", "Mushrooms", "Nicotine", "Semer", "VSA"
)

# Mapping of drug consumption classes to their meanings
consumption_mapping <- c(
  "CL0" = "Never Used",
  "CL1" = "Used over a Decade Ago",
  "CL2" = "Used in Last Decade", 
  "CL3" = "Used in Last Year",
  "CL4" = "Used in Last Month",
  "CL5" = "Used in Last Week",
  "CL6" = "Used in Last Day"
)

# Map Age values to their meaning
age_mapping <- c(
  "-0.95197" = "18-24",
  "-0.07854" = "25-34",
  "0.49788" = "35-44",
  "1.09449" = "45-54",
  "1.82213" = "55-64",
  "2.59171" = "65+"
)

# Map Gender values to their meaning
gender_mapping <- c(
  "0.48246" = "Female",
  "-0.48246" = "Male"
)

# Map Education values to their meaning
education_mapping <- c(
  "-2.43591" = "Left school before 16 years",
  "-1.73790" = "Left school at 16 years",
  "-1.43719" = "Left school at 17 years",
  "-1.22751" = "Left school at 18 years",
  "-0.61113" = "Some college or university, no certificate or degree",
  "-0.05921" = "Professional certificate/diploma",
  "0.45468" = "University degree",
  "1.16365" = "Masters degree",
  "1.98437" = "Doctorate degree"
)

# Map Country values to their meaning
country_mapping <- c(
  "-0.09765" = "Australia",
  "0.24923" = "Canada",
  "-0.46841" = "New Zealand",
  "-0.28519" = "Other",
  "0.21128" = "Republic of Ireland",
  "0.96082" = "UK",
  "-0.57009" = "USA"
)

# Map Ethnicity values to their meaning
ethnicity_mapping <- c(
  "-0.50212" = "Asian",
  "-1.10702" = "Black",
  "1.90725" = "Mixed-Black/Asian",
  "0.12600" = "Mixed-White/Asian",
  "-0.22166" = "Mixed-White/Black",
  "0.11440" = "Other",
  "-0.31685" = "White"
)
```

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Data Processing
###############################################################################

# Rename the columns 
colnames(drug_data) <- column_names

# Convert demographic columns to descriptive values
drug_data$Age <- age_mapping[as.character(drug_data$Age)]
drug_data$Gender <- gender_mapping[as.character(drug_data$Gender)]
drug_data$Education <- education_mapping[as.character(drug_data$Education)]
drug_data$Country <- country_mapping[as.character(drug_data$Country)]
drug_data$Ethnicity <- ethnicity_mapping[as.character(drug_data$Ethnicity)]

# Convert all drug consumption columns to descriptive values
for (col in drug_columns) {
  drug_data[[col]] <- consumption_mapping[as.character(drug_data[[col]])]
}
```

## Investigating missing values

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Data Cleaning - Missing values
###############################################################################

# Remove unnecessary column
drug_data <- drug_data[, -which(names(drug_data) == "ID")]

# Check for NA values in each column
na_by_column <- sapply(drug_data, function(x) sum(is.na(x)))
cat("NA values by column:")

# Print only columns with NA values
print(na_by_column[na_by_column > 0])  
```

Only two columns contain missing values, affecting approximately 5% of the 1885 observations. Given the nature of these variables and the completeness of the rest of the data, we assume participants deliberately withheld this information. Therefore, we replaced the missing values with "Not Provided", allowing us to treat these instances as a distinct category.

```{r echo=FALSE, cache=TRUE}
# Replace NA values with "Not Provided"
drug_data$Education[is.na(drug_data$Education)] <- "Not Provided"
drug_data$Ethnicity[is.na(drug_data$Ethnicity)] <- "Not Provided"

# Save the updated dataframe back to CSV
write.csv(drug_data, "Data/cleaned.csv")
```

## Investigating outliers

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Data Cleaning - Looking for outliers
###############################################################################
# Define numeric columns for outlier analysis
numeric_cols <- c("Nscore", "Escore", "Oscore", "Ascore", "Cscore", "Impulsive", "SS")

# Function to identify outliers using IQR method
identify_outliers_iqr <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  
  return(data.frame(
    min = min(x, na.rm = TRUE),
    q1 = q1,
    median = median(x, na.rm = TRUE),
    mean = mean(x, na.rm = TRUE),
    q3 = q3,
    max = max(x, na.rm = TRUE),
    iqr = iqr,
    lower_bound = lower_bound,
    upper_bound = upper_bound,
    n_outliers_below = sum(x < lower_bound, na.rm = TRUE),
    n_outliers_above = sum(x > upper_bound, na.rm = TRUE),
    total_outliers = sum(x < lower_bound | x > upper_bound, na.rm = TRUE),
    outlier_percentage = round(100 * sum(x < lower_bound | x > upper_bound, na.rm = TRUE) / length(x[!is.na(x)]), 2)
  ))
}

# Apply outlier detection to all numeric columns
outlier_summary <- data.frame()
for (col in numeric_cols) {
  result <- identify_outliers_iqr(drug_data[[col]])
  result$variable <- col
  outlier_summary <- rbind(outlier_summary, result)
}

# Create a function to visualize outliers with boxplots
plot_outliers <- function(drug_data, columns) {
  # Explicitly use reshape2::melt to avoid namespace issues
  melted_data <- reshape2::melt(drug_data[, columns], id.vars = NULL)
  
  # Create boxplot
  ggplot(melted_data, aes(x = variable, y = value)) +
    geom_boxplot(outlier.colour = "red", outlier.shape = 16, outlier.size = 2) +
    theme_minimal() +
    labs(title = "Boxplots with Outliers Highlighted",
         x = "Variable",
         y = "Value") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Visualize outliers
plot_outliers(drug_data, numeric_cols)
```

As can be seen from the box plots our data set has some values that are outside of the upper and lower bounds. All thought these values are technically outliers they are not extreme, still fall inside of the range of our expected values and conforms to a normal distribution.

# Exploratory Data Analysis

## Correlation Between Behavioral Measures

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Heatmap
###############################################################################

# Calculate the correlation matrix
cor_matrix <- cor(drug_data[numeric_cols], use = "pairwise.complete.obs")

# Convert the correlation matrix to a data frame for ggplot
cor_df <- melt(cor_matrix)
names(cor_df) <- c("Var1", "Var2", "Correlation")

# Create a ggplot2 correlation heatmap
ggplot(data = cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_fixed() +
  labs(
    title = "Correlation Matrix Behavioral Measures",
    x = "",
    y = ""
  )
```

The correlation matrix shows that certain personality traits tend to cluster thougher for example SS (Sensation Seeking) has a positive correlation with Escore (Extraversion), Oscore (Openness) and Impulsive while they in turn also have positive correlations which other and a negative correlation to Cscore (Conscientiousness) and Ascore (Agreeableness) while they have a positive correlation with each other.


## Comparing Behavioral Measure For Gender

```{r echo=FALSE, cache=TRUE}
###############################################################################
# Gender comparison
###############################################################################

# Calculating the means of the behavioral scores
gender_results <- data.frame(
  Trait = character(),
  Female_Mean = numeric(),
  Male_Mean = numeric(),
  stringsAsFactors = FALSE
)

for (col in numeric_cols) {
  # Calculate means only
  female_mean <- mean(drug_data[drug_data$Gender == "Female", col], na.rm = TRUE)
  male_mean <- mean(drug_data[drug_data$Gender == "Male", col], na.rm = TRUE)
  
  # Add to results dataframe with only needed values
  gender_results <- rbind(gender_results, data.frame(
    Trait = col,
    Female_Mean = female_mean,
    Male_Mean = male_mean,
    stringsAsFactors = FALSE
  ))
}

# Create readable trait names
gender_results$Trait_Name <- case_when(
  gender_results$Trait == "Nscore" ~ "Neuroticism",
  gender_results$Trait == "Escore" ~ "Extraversion",
  gender_results$Trait == "Oscore" ~ "Openness",
  gender_results$Trait == "Ascore" ~ "Agreeableness",
  gender_results$Trait == "Cscore" ~ "Conscientiousness",
  gender_results$Trait == "Impulsive" ~ "Impulsivity",
  gender_results$Trait == "SS" ~ "Sensation Seeking",
  TRUE ~ gender_results$Trait
)

# Create the plot directly from the results
ggplot(gender_results, aes(x = Trait_Name)) +
  geom_bar(aes(y = Female_Mean, fill = "Female"), stat = "identity", position = "dodge", width = 0.7, alpha = 0.7) +
  geom_bar(aes(y = Male_Mean, fill = "Male"), stat = "identity", position = position_dodge(width = 0.7), width = 0.7, alpha = 0.7) +
  scale_fill_manual(values = c("Female" = "#FF9999", "Male" = "#6699CC"),
                    name = "Gender") +
  labs(title = "Gender Differences in Behavioral Measures",
       x = "",
       y = "Mean Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        plot.title = element_text(hjust = 0.5, size = 14),
        legend.position = "top") +
  # Fixed axis limits from -0.25 to 0.25
  scale_y_continuous(limits = c(-0.25, 0.25)) +
  coord_flip()
```


The mean of all the Behavioral Measures is 0, the chart show the mean score broken down by gender for each Behavioral Measures. That chart shows that males tend to be more sensation seeking and impulsive but also more open, where females tend to be more impulsive but also more agreeable and conscientious.

## Comparing Education Level with Behavioral Measures


```{r, fig.width=7.5, fig.height=8.5, dev="pdf", echo=FALSE, cache=TRUE}
###############################################################################
# Education comparison
###############################################################################
# Order education levels
education_order <- c(
  "Left school before 16 years",
  "Left school at 16 years",
  "Left school at 17 years",
  "Left school at 18 years",
  "Some college or university, no certificate or degree",
  "Professional certificate/diploma",
  "University degree",
  "Masters degree",
  "Doctorate degree",
  "Not Provided"
)

# Calculate means by education level
education_means <- drug_data %>%
  group_by(Education) %>%
  summarize(
    n = n(),
    across(all_of(numeric_cols), ~mean(.x, na.rm = TRUE))
  ) %>%
  arrange(match(Education, education_order))

# Create a clean table view of the data
education_display <- education_means %>%
  select(-n) %>%
  rename(
    "Neuroticism" = Nscore,
    "Extraversion" = Escore,
    "Openness" = Oscore,
    "Agreeableness" = Ascore,
    "Conscientiousness" = Cscore,
    "Impulsivity" = Impulsive,
    "Sensation Seeking" = SS
  )

# Create a data frame for plotting
education_plot_data <- education_means %>%
  select(-n)

# Simplified visualization approach with more space for the title
# Set up the plotting parameters
par(mfrow = c(4, 2), mar = c(8, 4, 2, 1), oma = c(0, 0, 3, 0))

# Define the trait columns and their display names
trait_cols <- c("Nscore", "Escore", "Oscore", "Ascore", "Cscore", "Impulsive", "SS")
trait_names <- c("Neuroticism", "Extraversion", "Openness", "Agreeableness", 
                "Conscientiousness", "Impulsivity", "Sensation Seeking")

# Plot each trait separately
for (i in 1:length(trait_cols)) {
  col <- trait_cols[i]
  trait_name <- trait_names[i]
  
  # Create a simple barplot
  bp <- barplot(education_plot_data[[col]], 
          names.arg = rep("", nrow(education_plot_data)),
          col = ifelse(education_plot_data[[col]] > 0, "salmon", "lightblue"),
          main = trait_name,
          border = NA,
          las = 2,
          ylim = c(min(education_plot_data[[col]]) - 0.05, 
                  max(education_plot_data[[col]]) + 0.05),
          cex.main = 0.9,
          cex.axis = 0.8)
  
  # Add a reference line at y=0
  abline(h = 0, lty = 2, col = "gray")
  
  # Add abbreviated education labels
  edu_labels <- c("Before 16", "At 16", "At 17", "At 18", "Some college", 
                 "Prof cert", "University", "Masters", "Doctorate", "Not Provided")
  edu_labels <- edu_labels[1:length(bp)]  # Ensure we have the right number of labels
  
  # Add text labels at the bottom
  text(bp, par("usr")[3] - 0.02, srt = 45, adj = 1, labels = edu_labels, 
       xpd = TRUE, cex = 0.7)
}

# Add a title for the entire set of plots with more space
mtext("Personality Traits by Education Level", side = 3, line = 1, 
      outer = TRUE, cex = 1.2, font = 2)
```
It is not very clear at first glance but when you study the table closely it becomes clear that traits that can be precived as bad like Neuroticism, Impulsivity and Sensation Seeking	is more prevelent with lower education levels inccluding Not Provided and steadily decrease as the level of education increases.  

## Analysis of Seremon Usage

```{r dev="pdf", echo=FALSE, cache=TRUE}
###############################################################################
# Seremon
###############################################################################

# Count Semeron users vs non-users
semeron_counts <- drug_data %>%
  group_by(Semer) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
```

```{r}
# Create a nicely formatted table for the detailed counts
semeron_table <- semeron_counts %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 2), "%")) %>%
  rename(`Usage Category` = Semer) %>%
  kable(caption = "Semeron Usage Categories", 
        booktabs = TRUE, 
        col.names = c("Usage Category", "Count", "Percentage"))

# For PDF output, apply specific styling
if(knitr::is_latex_output()) {
  semeron_table <- semeron_table %>%
    kable_styling(latex_options = c("striped", "hold_position"), 
                  full_width = FALSE,
                  position = "center") %>%
    row_spec(0, bold = TRUE) %>%
    column_spec(1, width = "5cm") %>%
    column_spec(2, width = "2cm") %>%
    column_spec(3, width = "2.5cm")
}

# Display the table
semeron_table
```
# Source
https://archive.ics.uci.edu/dataset/373/drug+consumption+quantified